\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}

\newcommand{\dIV}{\Delta\mathrm{IV}}

\title{Short-Horizon Forecasting of the SPY Implied-Volatility Surface\\
  \large A pre-registered study on Databento OPRA data}
\author{Viranca R.I. Balsingh\thanks{Forecasting results correspond to
\texttt{qfdev\_research} commit \texttt{3eba5bf8}; the execution-replay
design of Section~\ref{sec:replay} is locked at commit \texttt{613a309f},
pushed before any replay computation; the single replay run used code at
\texttt{884ff426} and its adjudication is committed at \texttt{78c035d4}.}\\
\texttt{virancab@gmail.com}}
\date{July 21, 2026}

\begin{document}
\maketitle

\begin{abstract}
Are short-horizon moves of the SPY implied-volatility surface predictable,
and does the forecast retain positive vega-attributed edge after observable
spread costs? I build one-minute, quality-gated and static-arbitrage-screened
volatility-surface states from OPRA consolidated top-of-book quotes
($4.4\times10^{10}$ records over the initial 21-session panel,
June~2--July~1, 2026, later extended to 64 sessions spanning
March~31--July~1), then forecast per-cell implied-volatility changes at
5--60-minute horizons under a purged walk-forward protocol with mechanical
leakage probes and shuffled-label negative controls. At the primary
15-minute horizon, ridge regression on a compact set of time-$t$ features
improves vega-weighted MSE over the persistence null by 16.7\% on the June
panel (25.9\% unweighted); on the expanded three-month panel its unweighted
skill is 26.1\%. A GPU sweep of temporal convolutional networks shows a deep
sequence model adds essentially nothing: within the tested configurations,
receptive fields beyond 15 minutes produced no additional out-of-sample
skill. Pre-registered feature ablations reveal that a finding as intuitive
as ``Greeks help'' flips sign between a one-month and a three-month panel.
A final symmetric experiment finds that, in the tested one-minute per-cell
representation, Lee--Ready-signed prints add no measurable incremental
out-of-sample skill beyond quote-based order-flow features at any horizon.
Every expectation was registered before the corresponding run; negative
results are reported as registered. The economic question is settled by a
pre-registered execution-aware replay (Section~\ref{sec:replay}) whose
design, endpoints, and verdict taxonomy were locked before any replay
computation and which ran exactly once: at the primary 15-minute horizon
the forecast does \emph{not} clear the touch (mean entry-cost markout
$-\$0.0030$ per share, 95\% day-bootstrap CI $[-0.0040, -0.0021]$), even
though it genuinely selects favorable repricing; a registered
attribution decomposition shows the model prices cell-median dislocations
that the tradable contract realizes only fractionally at short horizons.
An exploratory horizon sweep, reported with its multiple-comparison
caveat, finds the same forecasts clear entry costs decisively at 60
minutes, naming the next pre-registerable hypothesis.
\end{abstract}

\section{Question and setup}
\label{sec:setup}

The object being forecast is the surface state
$\mathrm{IV}_t(\text{cell})$: for each cell of a fixed
(days-to-expiry $\times$ log-moneyness) grid, the target is the forward
change
\[
  \dIV(t, h) \;=\; \mathrm{IV}(t+h) - \mathrm{IV}(t),
  \qquad h \in \{5, 15, 30, 60\}\ \text{minutes},
\]
with 15 minutes as the primary horizon. The null model is persistence
($\dIV = 0$), the surface staying where it is, which at these horizons is
a genuinely hard baseline, not a strawman.

``Good'' is defined by what a desk would do with the forecast, not by raw
RMSE. The eventual use is relative-value: to first order, the
volatility-attributed premium change of an option is approximately
$\text{vega}_t \times \dIV$ per share (dollar results per standard contract
multiply quoted premium changes by the 100-share multiplier). This
first-order attribution excludes delta, gamma, theta, and volga effects,
hedging, and financing; the execution replay of Section~\ref{sec:replay}
therefore reports a \emph{vega-linearized} edge, never total option P\&L.
The headline forecasting metric is \emph{vega-weighted} MSE skill versus
persistence (defined exactly in Section~\ref{sec:method}), alongside the
per-minute cross-sectional rank information coefficient. The tradeable scope
was locked in advance to short-dated wings (expiries within seven days, strikes off the immediate at-the-money point) because that is where both
the movement and a realistic edge concentrate.

\section{Data and surface construction}
\label{sec:data}

The quote source is Databento OPRA consolidated NBBO top-of-book
(\texttt{cmbp-1}) for all listed SPY options, with Nasdaq \texttt{mbp-1} for
the underlying. Three nested samples appear in this memo:

\begin{table}[h]
\centering
\small
\begin{tabular}{llcl}
\toprule
Stage & Dates (2026) & Sessions & Purpose \\
\midrule
Initial panel & Jun 2 -- Jul 1 & 21 & Baselines and first ablations \\
Expanded panel & Mar 31 -- Jul 1 & 64 & Architecture sweep; pre-registered replications \\
Replay prediction dump & Apr 15 -- Jul 1 & 54 test days & Execution-aware replay (Sec.~\ref{sec:replay}) \\
\bottomrule
\end{tabular}
\caption{The three sample stages. The replay dump contains the 7{,}633{,}773
out-of-sample prediction rows produced by the walk-forward protocol on the
expanded panel; its 54 test days are the expanded panel's out-of-sample
days.}
\label{tab:samples}
\end{table}

The initial 21-session panel (June~2 to July~1; ``the June panel'' below)
covers roughly 858\,GB of raw quotes, 44.05 billion records, verified
row-for-row against the vendor's manifest.

Quotes are compacted into one-minute surface snapshots. Each snapshot fits
SVI slices \cite{gatheral2014} per expiry with quality gating (85.7\% of
slices trusted on the reference validation day; the rest fall back to direct
interpolation), recovers forwards from put--call parity (100\% coverage),
and writes IV, Black-76 \cite{black1976} Greeks, and quality flags onto a
fixed grid of 9 DTE buckets $\times$ 14 log-moneyness nodes: 126 cells
per minute. The June panel used for forecasting is 886{,}467 cell-minutes.

Every session must pass an automated validation gate before it enters the
panel. Among contract-minutes passing the quote-eligibility filters (finite
two-sided quotes with $0 \le \text{bid} \le \text{ask}$ and a positive mid, the builder's \texttt{ok} flag), the fraction with a solvable IV must be
$\geq 90\%$; coverage fractions quoted against \emph{all} raw
contract-minutes, including ineligible and malformed quotes, are lower by
construction and are not the gated quantity. Further gates: median put--call
IV disagreement $\leq 1.5$ vol-points, butterfly-negativity
(a static-arbitrage smell) below threshold, SVI fit RMSE bounds, and an
external sanity check of ATM IV against an independent reference within
3 vol-points. All 21 June sessions passed.

\section{Methodology and pre-registration}
\label{sec:method}

\paragraph{Models.} Five model classes appear throughout: persistence (the
$\dIV=0$ null); a constant train-mean baseline (the drift-aware null used by
the negative controls); ridge regression ($\alpha = 1$ on standardized
features) as the locked reference model; LightGBM as a vega-weighted
challenger; and a causal dilated temporal convolutional network (TCN)
\cite{bai2018} trained with a vega-weighted Huber loss.

\paragraph{Vega-weighted metric.} The headline metric is skill versus
persistence under vega-weighted MSE,
\[
  \mathrm{MSE}_v \;=\; \frac{\sum_i w_i\,(y_i - \hat y_i)^2}{\sum_i w_i},
  \qquad w_i = \text{vega}_i,
\]
where $\text{vega}_i$ is the raw per-cell Black-76 vega written by the
surface builder at decision time $t$ (no cap, floor, or normalization; rows
with missing vega are excluded), and skill is
$1 - \mathrm{MSE}_{\text{model}} / \mathrm{MSE}_{\text{persistence}}$
computed under the same weights. Weighting always enters
\emph{evaluation}; it enters \emph{training} only where stated: LightGBM
via per-row sample weights and the TCN via its vega-weighted loss; ridge is
trained unweighted.

\paragraph{Cross-sectional rank IC.} At each minute with at least 20
in-scope cells, the Spearman rank correlation between predicted and realized
$\dIV$ is computed across cells; the reported figure is the unweighted mean
over minutes. Because 15-minute labels are sampled every minute, the raw
$t$-statistics over minutes are overlap-inflated; deflating by $\sqrt{15}$
is quoted alongside where it matters.

\paragraph{Validation protocol.} Five-fold expanding-window walk-forward
over the minute axis, purged and embargoed by the longest horizon
(60 minutes) so overlapping labels cannot leak across the train/test
boundary \cite{lopezdeprado2018}; roughly 600--700k out-of-sample rows per
horizon on June alone. Feature standardization is fit on training folds
only.

\paragraph{Leakage controls.} Two mechanical probes run against the feature
builder itself: truncation invariance (features computed on a panel cut at
time $t$ must be bitwise identical to the full-panel build) and
future-input perturbation (corrupting all raw inputs after $t$ must leave
features at or before $t$ unchanged). The probes were validated by injecting
a deliberate future-IV feature, which they caught. In addition,
shuffled-label negative controls were run for every model class: within each
walk-forward fold, \emph{training} labels are permuted after the
purge/embargo (test labels intact) and the models retrained. Because the
target has non-zero mean over long windows, shuffled models are judged
against the constant train-mean baseline rather than the zero null; otherwise target drift masquerades as leakage. Measured feature-driven
excess collapses to $\leq +0.004$ against a $0.02$ tolerance (versus
$+0.17$ to $+0.31$ on real labels), and feature importances go flat.

\paragraph{Significance tests.} Reported Diebold--Mariano
\cite{diebold1995} $p$-values test the per-minute loss-differential series
between two arms scored on identical row sets (cross-sectional losses
aggregated within the metric's weighting), with a Newey--West long-run
variance at lag equal to the horizon length in minutes to absorb label
overlap. They are computed per (model, horizon, metric) cell and are
\emph{not} adjusted for multiplicity across the grid; they serve as
supporting diagnostics. The primary evidence standard everywhere is the
registered effect threshold with the sign holding in every walk-forward fold
and every calendar month.

\paragraph{Pre-registration.} Before each experiment round, the expected
outcome, the primary endpoints, and the decision rules were written down and
committed. Example: Greeks become default-on for a model class only if they
add at least $+1.0$ percentage point of vega-weighted skill on at least two
horizons \emph{with the sign holding in every walk-forward fold}. The point
is to make the negative results as legible as the positive ones.

\section{Results}
\label{sec:results}

\subsection{Baselines beat persistence; the signal is mostly linear}

Walk-forward skill versus persistence on the June panel:

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
Skill vs.\ persistence & 5m & 15m & 30m & 60m \\
\midrule
Ridge, plain MSE        & +25.2\% & +25.9\% & +24.6\% & +30.9\% \\
LightGBM, plain MSE     & +21.5\% & +16.7\% & +18.3\% & +24.9\% \\
\addlinespace
Ridge, vega-weighted    & +19.6\% & +16.7\% & +8.3\%  & +8.0\%  \\
LightGBM, vega-weighted & +17.8\% & +11.9\% & +7.0\%  & +1.3\%  \\
\bottomrule
\end{tabular}
\caption{Out-of-sample skill, 21-session June panel, 5-fold purged
walk-forward. LightGBM rows use vega sample-weighting in training. Values
are point estimates; day-clustered uncertainty is currently reported only
for the registered adjudications (fold ranges, DM diagnostics) and the
replay's day bootstrap; extending day-bootstrap intervals to this table
is queued work (Section~\ref{sec:limits}).}
\label{tab:june}
\end{table}

Ridge matches or beats LightGBM on plain MSE at every horizon: the signal is
largely linear in this feature set. The weighting matters more than the
model: an unweighted LightGBM is \emph{worse than persistence} at the
15-minute horizon once evaluated vega-weighted ($-1.3\%$), and simply
training with vega sample-weights swings it to $+11.9\%$, a 13-point difference from an evaluation choice, before any modeling work.
Directional accuracy decays from 58\% at 5 minutes to 53\% at 60 minutes.
On short-DTE wings, the locked reference model achieves a mean per-minute
cross-sectional rank IC of $+0.40$ (raw $t\approx116$ over minutes,
$\approx 30$ after $\sqrt{15}$ overlap deflation; 91\% of minutes positive)
with a vega-weighted calibration slope of $0.992$ (vega-WLS of realized on
predicted through the origin), so predicted magnitudes can be taken at face
value in the tradeable scope. The adjacency of surface cells makes the
per-minute cross-section dependent, so the IC's minute count overstates its
effective sample size; the fold- and month-consistency standards of
Section~\ref{sec:method} are the load-bearing evidence.

\subsection{A negative result on architecture}

An 11-run TCN sweep on the full 64-session span (single RTX~4060~Ti,
$\sim$15h wall) tested whether a deep sequence model finds anything the
cheap models miss. It does not: TCN skill ties ridge and LightGBM at every
horizon: at 15 minutes the \emph{plain-MSE} skill fractions are ridge
$0.261$, LightGBM $0.251$, TCN $0.248$ (the sweep's summary metric is
unweighted skill versus persistence). A receptive-field ablation is the
sharpest summary: growing the TCN's context from 3 to 7 minutes helps, and
within the tested configurations, increasing the receptive field beyond
15 minutes produced no additional out-of-sample skill. That bounds the
incremental value of longer memory for this model class and representation;
it does not establish that no longer-memory information exists. In a
capacity sweep, additional dropout did not improve out-of-sample performance
(dropout $0$ was optimal), suggesting insufficient regularization was not
the limiting factor in the tested capacity range. The feature set does the
work, not the architecture; the next lever is cross-sectional structure, not
a bigger network.

\subsection{Feature ablations: importance is not generalization}
\label{sec:ablations}

On June, adding Black-76 Greeks as features \emph{hurt} LightGBM by 4 to 11
skill points depending on horizon (worst at 30m: $-10.6$pp), while leaving
ridge indifferent, even though Greeks captured 8.6\% of the model's split
gain at that same horizon. A plausible interpretation was that gamma, theta
and vega are near-deterministic transforms of features already present, and
that the redundancy encouraged unstable tree splits that spend themselves on
out-of-sample variance. The expanded-panel replication in
Section~\ref{sec:replication} was designed to test exactly that
interpretation. Quote-based microstructure features (order-flow imbalance,
quote imbalance) added $+1.5$pp for LightGBM at 30/60m and nothing below
that. Both findings were treated as hypotheses to re-test, not conclusions.

\section{What changed between rounds, and why}
\label{sec:replication}

The June ablations were re-run as a pre-registered $2\times2$
(Greeks $\times$ microstructure) on the three-month panel (the five model classes of Section~\ref{sec:method} $\times$ 4 horizons $\times$ 4 arms), with predictions and decision rules committed before launch. The registered
expectations were: ridge flat under Greeks ($|\Delta| < 0.5$pp), LightGBM
hurt by $\geq 1$pp on $\geq 2$ horizons (replicating June), TCN
approximately flat, and no Greeks$\times$micro interaction.

The replication broke the June story in an instructive way. All contrasts
below are computed on identical row sets, with per-fold and per-calendar-month
deltas and Diebold--Mariano tests as described in
Section~\ref{sec:method}. The LightGBM
Greeks harm, the largest and most intuitive June effect, did
\emph{not} replicate: at the 30-minute replication target the sign flipped
from $-10.6$pp to $+0.7$pp (all five folds positive, DM $p<10^{-4}$), with
$+0.3$ to $+0.7$pp across the other horizons. Ridge, registered as flat,
instead gained $+1.9$pp of vega-weighted skill at 30 minutes and $+4.8$pp at
60 from Greeks, with the sign holding in every walk-forward fold and every
calendar month, so under the rule fixed in advance, Greeks flipped
default-on for the ridge class (and only there: no other model class met the
threshold; a fold-consistent TCN cell at 30 minutes had no vega-weighted
counterpart and was discarded under the registered stability standard). The
registered mechanism story, that the June harm was specific to greedy split selection, failed its own preconditions and is reported as wrong.
The microstructure edge shrank from $+1.5$pp to $\sim$$+0.3$pp at 30/60
minutes; its registered replication target (LightGBM 60m $\Delta > 0$)
formally held ($+0.31$pp, all folds positive, DM $p=0.026$) but attenuated
fivefold. At the 15-minute story horizon every effect was null. The plain
reading: one month is one regime, and feature-ablation signs measured on it
do not transfer. The pre-registration is what makes this a finding rather
than an embarrassment: the June conclusions were recorded, the re-test
criteria were fixed in advance, and the reversal is reportable exactly
because nothing was re-fit after seeing the answer.

\subsection*{A symmetric test of trade flow}

The quote-based order-flow features raise an obvious question: do actual
\emph{prints} carry anything the book does not? A second pre-registered
$2\times2$ (microstructure $\times$ signed trade flow) added Lee--Ready
\cite{leeready1991} signed volume per cell from OPRA trades, engineered with
exactly the same
30-minute temporal treatment as the quote-pressure features so the
comparison could not be confounded by representation, and framed two
competing hypotheses in advance: prints add information beyond posted
pressure (H-new) versus quotes impound trades within the minute (H-span).
Results consistently favored H-span over H-new within the registered
representation. The primary endpoint (LightGBM skill change from flow given the quote features, 60-minute horizon) was $-0.07$pp
(DM $p=0.46$); the largest ridge cell was $+0.07$pp; and the substitution
contrast (flow alone versus neither) was equally null, so there was no
flow credit for the quote features to steal. Two TCN cells produced large
point estimates ($+3.0$pp at 30m vega-weighted, DM $p=0.004$) with per-fold
deltas spanning $-1.1$ to $+11.5$pp, exactly the multiple-comparisons
bait the registered fold-consistency standard exists to discard. Two
caveats are owned rather than hidden: Lee--Ready misclassifies option
trades (multi-leg packages especially), and per-cell signed volume
conflates directional with volatility demand; the strongest untested
variant, chain-level vega-weighted net flow, is named in
Section~\ref{sec:limits} rather than quietly run.

The replication round also doubled as an engineering exhibit: the two
overlapping arms were recomputed on a second machine (Windows, pinned to
the same library versions) and all 64 recomputed ridge and LightGBM summary
values matched the Linux results bit-for-bit. (The claim is scoped to those
table values; row-level prediction files were not hashed.)

\section{Execution-aware replay: does the edge survive the spread?}
\label{sec:replay}

Vega-weighted forecast skill is not a trading result. This section closes
that gap with a pre-registered execution replay that maps each in-scope cell
forecast to an actual quoted contract and charges observable spread costs.
The full design, endpoints, falsifiable expectation bands, and verdict
taxonomy below were committed (\texttt{613a309f}) \emph{before any replay
computation existed}; the replay runs exactly once against them.

\subsection{Registered design (summary)}

For each (minute, cell) a single representative OTM-side contract is chosen
on time-$t$ information only (nearest log-moneyness to the node among
quote-valid contracts, tie-broken by relative spread); the exit is the
\emph{same} contract at the exact quote $h$ minutes later, with no forward
fill; a missing exit is counted as unfillable and excluded from markout,
never silently filled. Vega is the per-contract Black-76 vega at $t$. A
trade fires iff $|\text{vega}_t \cdot \lambda \cdot \widehat{\dIV}| >
\theta \cdot \text{half-spread}_t + s$, with $\theta = 1$, $\lambda = 1$,
$s = 0$ primary and $\theta \in \{1, 1.5, 2\}$, slippage
$s \in \{0, 0.005, 0.01\}$ \$/share, and a walk-forward-calibrated
$\lambda$ arm as sensitivities. Three outcome quantities are kept distinct:
the gross vega-linearized edge
$\mathrm{sign}(\widehat{\dIV})\cdot\text{vega}_t\cdot(\mathrm{IV}_{t+h} -
\mathrm{IV}_t)$; the \emph{entry-cost-adjusted markout} (gross minus
half-spread at entry, marked to $t{+}h$ mid), the statistical primary;
and the \emph{round-trip-touch net} (additionally minus the exit
half-spread), which drives the economic verdict. Scope is the registered
tradeable region (DTE $\leq 7$, off-ATM); out-of-scope cells are evaluated
hypothetically and reported separately. Inference is a day-level bootstrap
($N = 10{,}000$) over the 54 test days. A greedy per-cell cooldown variant
provides a non-overlapping replay; its cumulative output is a cumulative
one-contract non-overlapping edge, \emph{not} a portfolio equity curve, and
is reported with concurrency diagnostics. A delayed-entry arm (decide at
$t$, enter at the $t{+}1$-minute touch) bounds the latency objection.

\subsection{Primary endpoint and decision rule}

\begin{quote}
\itshape The registered primary endpoint is the pooled mean
entry-cost-adjusted vega-linearized markout per fillable gated trade (ridge, 15-minute horizon, in-scope, $\theta = 1$, $\lambda = 1$, $s = 0$) with a 95\% day-bootstrap confidence interval.
\end{quote}

The endpoint is invalidated (reported as ``diagnostics only'', no headline
claim) if gated exit unfillability exceeds 5\% overall or is more than twice
the overall rate in the top predicted-edge decile; selective
unfillability where the model claims the most edge would make the markout
uninterpretable.

\subsection{Registered verdict taxonomy}

The prose conclusion is constrained to one of three registered verdicts:

\begin{itemize}
  \item \textbf{Survives observable entry costs.} Primary CI entirely above
    zero, round-trip-touch point estimate $\geq 0$ with its sign preserved
    in the non-overlapping variant, validity condition intact.
  \item \textbf{Markout survives entry costs; round-trip execution not
    demonstrated.} Primary CI above zero, but the round-trip criterion
    fails.
  \item \textbf{Does not survive costs.} Primary CI straddling or below
    zero, or the endpoint invalidated by selective missing exits, reported as such: 15-minute surface forecasts of this quality do not
    clear the touch.
\end{itemize}

Six falsifiable expectation bands (gate pass rate, gross selection
monotonicity in $\theta$, the honest spread-survival band, LightGBM
overconfidence as a trading statement, out-of-scope deterioration, and
non-overlap sign preservation) were registered alongside; the strong result (round-trip positive with CI above zero) is registered as a
\emph{surprise}, not the expectation.

\subsection{Results}
\label{sec:replay-results}

One amendment was registered between the design commit and the run: the
registered one-day smoke test fired its attribution gate (on the smoke day the tradable contract's $\dIV$ realized only ${\sim}0.2\times$ the
cell-median target the model predicts), so a pre-specified attribution
decomposition (floating cell median vs.\ fixed-membership median vs.\
single contract) was added to the endpoint set (\texttt{884ff426})
\emph{before the full run}, with the primary endpoint and every expectation
band unchanged. The replay then ran exactly once: 54 test days, three
models, all registered arms.

\paragraph{Gates and validity.} The mechanical self-test and both lookahead
probes pass. One registered bound breached marginally and is disclosed
rather than amended: 2.4--2.6\% of predictions per horizon matched no
tradable contract, against a registered 2\% bound. Every such case is a
far-wing cell whose node contained no quote-valid OTM-side contract, a
cell on which the replay's instrument is not defined; there were zero join
failures. The validity condition holds on both clauses: gated exit
unfillability is 1.53\% ($<5\%$) and \emph{falls} with predicted edge
(0.37\% in the top decile), so missing exits do not concentrate where the
model claims the most edge.

\paragraph{Primary endpoint.} Ridge, 15-minute horizon, in scope,
$\theta=1$, $\lambda=1$, $s=0$: mean entry-cost-adjusted markout per
fillable trade $\mathbf{-\$0.0030}$ per share, 95\% day-bootstrap CI
$[-0.0040,\,-0.0021]$, over 509{,}946 trades on 54 of 54 days (the
equal-day-weighted mean, $-\$0.0029$, agrees); round-trip-touch net
$-\$0.0098$ $[-0.0108,\,-0.0088]$. The verdict is the registered third:
\textbf{15-minute surface forecasts of this quality do not survive costs: they do not clear the touch.}

\paragraph{Registered expectations.} Of the six bands, three held and three
were rejected, and the rejections are the informative part.
\emph{Supported:} gross selection works (conditional mean gross
vega-linearized edge $+0.0033$ at $\theta=1$, rising monotonically to
$+0.0040$ at $\theta=2$, same pattern for all three models; the failure is cost-sized, not signal-sized); out-of-scope cells are worse for the
registered model (conditional gross $+0.0023$ vs.\ $+0.0033$ in scope);
the non-overlapping replay preserves every sign (trade-count ratio 0.086,
primary $-0.0030 \to -0.0038$). \emph{Rejected:} the gate pass rate came
out at 71.8\% against a registered $[2, 30]\%$ band: the gate believes
the cell-median move, while the fill gets the contract (the mechanism the
decomposition below quantifies); LightGBM did \emph{not} over-trade
relative to ridge (pass rate 55.9\%, per-trade markout $-0.0028$); and the
markout did not survive entry costs (the primary above).

\paragraph{Attribution: what the model predicts vs.\ what a contract
realizes.} The amended decomposition separates membership churn from
single-contract attenuation
(Table~\ref{tab:attribution}). Membership churn is a roughly constant
haircut: the fixed-membership cell median tracks the target at
corr${}\approx 0.7$ at every horizon. The horizon-dependent loss is
per-contract: at 5--15 minutes the cell-median $\dIV$ the model forecasts
is dominated by quote-bounce of the cell's wide members; that bounce is
mean-reverting and genuinely \emph{predictable}, but not realizable on the tight tradable
contract (regression slope of contract on fixed-median $\dIV$: 0.09 at 5m,
0.25 at 15m). By 60 minutes the contract realizes roughly $0.7\times$ the
cell move.

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
$h$ & corr$(y, \text{fixed})$ & corr$(\text{fixed}, \text{contract})$
  & slope (contract on fixed) & corr$(y, \text{contract})$ \\
\midrule
5m  & 0.74 & 0.28 & 0.09 & 0.13 \\
15m & 0.70 & 0.47 & 0.25 & 0.26 \\
30m & 0.70 & 0.64 & 0.44 & 0.41 \\
60m & 0.73 & 0.81 & 0.68 & 0.60 \\
\bottomrule
\end{tabular}
\caption{Registered attribution decomposition, in scope. $y$ is the
floating cell-median target the model is trained on; ``fixed'' freezes the
cell's membership at $t$; ``contract'' is the single tradable contract the
replay fills. Cell-level predictability survives at all horizons; its
tradable share grows with horizon.}
\label{tab:attribution}
\end{table}

\paragraph{Exploratory: the economics are monotone in horizon.}
\begin{sloppypar}
With the
explicit multiple-comparison caveat that the registered primary was 15
minutes, the per-trade economics improve with horizon exactly as the
decomposition predicts: entry-cost markout $-0.0058$ at 5m, $-0.0030$ at
15m, $+0.0009$ at 30m (CI straddling zero), and $\mathbf{+0.0083}$ at 60m
with CI $[+0.0043, +0.0117]$ entirely positive;
the 60m round-trip-touch net is $+0.0015$ $[-0.0026, +0.0049]$: breakeven, not a win. The 60m entry-cost result survives the registered
honesty arms: $+0.0078$ under one-minute delayed entry and $+0.0077$ under
non-overlapping sampling; on the slippage grid it remains positive through
$s = \$0.005$, whereas at 15m the same $s$ is immediately fatal
($-0.0076$). This is reported as what it is: not a claimed result, but the
precise, pre-registerable hypothesis the registered run points to, namely
to confirm the 60-minute horizon as primary on new data.
\end{sloppypar}

\section{Limitations and next steps}
\label{sec:limits}

\begin{itemize}
  \item \textbf{One underlier, one quarter.} Everything here is SPY,
    March~31 -- July~1, 2026 (approximately Q2). The regime-sensitivity
    result in Section~\ref{sec:replication} is itself the argument for not
    extrapolating further than the data supports.
  \item \textbf{Execution-aware markout, not a portfolio backtest.} The
    replay of Section~\ref{sec:replay} maps cell forecasts to actual option
    quotes and charges observable spread costs, but its primary outcome is
    a vega-linearized markout. It does not model delta hedging, nonlinear
    Greeks, queue position, displayed-size constraints, market impact,
    portfolio capital, or asynchronous latency (the delayed-entry arm
    bounds, but does not remove, the latency objection). The same-bar touch
    assumption is an execution benchmark, not a live-fill claim. Source
    quotes may have been forward-filled by up to five minutes in the
    surface builder, and the slim contract extract does not carry a per-row
    staleness flag.
  \item \textbf{Point estimates in the forecasting tables.} The headline
    skill tables report point estimates; day-clustered uncertainty
    currently exists only where the registered adjudications required it
    (fold ranges, per-month splits, DM diagnostics) and in the replay's day
    bootstrap. Extending day-bootstrap intervals to
    Table~\ref{tab:june} is queued.
  \item \textbf{Surface gates are static-arb smells, not proofs.} The
    validator checks butterfly negativity and put--call consistency;
    a full calendar/butterfly no-arbitrage repair across the grid is open
    work.
  \item \textbf{The flow null is a null for one representation.} Per-cell,
    one-minute, Lee--Ready-signed volume with window aggregates adds
    nothing; that bounds the \emph{average} contribution, not a localized
    one. The named untested variants: chain-level vega-weighted net flow
    (attacks the dealer-inventory channel directly, endpoint: LightGBM
    vega-weighted skill at 30/60m), skill conditioned on cell-minutes with
    nonzero window flow (attacks sparsity dilution), and multi-leg print
    exclusion (attacks sign misclassification).
  \item \textbf{Next:} pre-register and confirm the 60-minute horizon as
    primary on new data (the one hypothesis the replay's exploratory sweep
    names; Section~\ref{sec:replay-results}); cross-sectional feature
    structure (the direction the architecture sweep points to); per-model
    feature defaults in code (the ridge/Greeks flip is adjudicated but
    ridge and LightGBM currently share one feature matrix).
\end{itemize}

\begin{thebibliography}{9}

\bibitem{black1976}
F.~Black.
\newblock The pricing of commodity contracts.
\newblock \emph{Journal of Financial Economics}, 3(1--2):167--179, 1976.

\bibitem{gatheral2014}
J.~Gatheral and A.~Jacquier.
\newblock Arbitrage-free SVI volatility surfaces.
\newblock \emph{Quantitative Finance}, 14(1):59--71, 2014.

\bibitem{leeready1991}
C.~M.~C. Lee and M.~J. Ready.
\newblock Inferring trade direction from intraday data.
\newblock \emph{Journal of Finance}, 46(2):733--746, 1991.

\bibitem{bai2018}
S.~Bai, J.~Z. Kolter, and V.~Koltun.
\newblock An empirical evaluation of generic convolutional and recurrent
  networks for sequence modeling.
\newblock arXiv:1803.01271, 2018.

\bibitem{diebold1995}
F.~X. Diebold and R.~S. Mariano.
\newblock Comparing predictive accuracy.
\newblock \emph{Journal of Business \& Economic Statistics}, 13(3):253--263,
  1995.

\bibitem{lopezdeprado2018}
M.~L\'opez~de~Prado.
\newblock \emph{Advances in Financial Machine Learning}.
\newblock Wiley, 2018.

\end{thebibliography}

\end{document}
